---
title: "2022_09_26_chapter4"
author: "Matthew Davis"
date: "2022-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 4 Subsetting

### 4.1 Introduction
`str()` shows the objects structure
`View()` is the interactive RStudio viewer

### 4.2 Selecting multiple elements
`[]` selects any number of elements from a vector
```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

Postive integers 
```{r}
# Positive integers return elements at the specified positions
x[c(3, 1)]

x[order(x)]

# Duplicate indices will duplicate values
x[c(1, 1)]

# Real numbers are silently truncated to integers
x[c(2.1, 2.9)]
```

Negative integers
```{r}
# Negative integers exclude elements at the specified positions
x[-c(3, 1)]

# You can't mix positive and negative integers in a single subset
x[c(-1, 2)]
```

Logical vectors
```{r}
#Logical vectors select elements where the corresponding logical value is TRUE. This is probably the most useful type of subsetting because you can write an expression that uses a logical vector:
x[c(TRUE, TRUE, FALSE, FALSE)]

x[x > 3]

```

Recycling rules
```{r}
# In x[y], what happens if x and y are different lengths? The behaviour is controlled by the recycling rules where the shorter of the two is recycled to the length of the longer. 

x[c(TRUE, FALSE)]
# Equivalent to
x[c(TRUE, FALSE, TRUE, FALSE)]
```

Nothing
```{r}
# Nothing returns the original vector. 
x[]
```

Zero
```{r}
# Zero returns a zero-length vector.
x[0]
```

Character vectors
```{r}
# If the vector is named, you can also use character vectors to return elements with matching names.
(y <- setNames(x, letters[1:4]))
y[c("d", "c", "a")]

# Like integer indices, you can repeat indices
y[c("a", "a", "a")]

# When subsetting with [, names are always matched exactly
z <- c(abc = 1, def = 2)
z[c("a", "d")]
```

Factors
```{r}
# Factors are not treated specially when subsetting. This means that subsetting will use the underlying integer vector, not the character levels. You should avoid subsetting with factors.
y[factor("b")]
```

You can subset higher-dimensional structures in three ways:

With multiple vectors.
With a single vector.
With a matrix.

```{r}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

a[c(TRUE, FALSE, TRUE), c("B", "A")]

a[0, -2]
```

`[]` simplifies the result to the lowest possible dimensionality
```{r}
a[1, ]
a[1,1]
```

Because both matrices and arrays are just vectors with special attributes, you can subset them with a single vector, as if they were a 1D vector.
```{r}
vals <- outer(1:5, 1:5, FUN = "paste", sep = ",")
vals

vals[c(4, 15)]
```

You can also subset higher-dimensional data structures with an integer matrix (or, if named, a character matrix). Each row in the matrix specifies the location of one value, and each column corresponds to a dimension in the array.
```{r}
select <- matrix(ncol = 2, byrow = TRUE, c(
  1, 1,
  3, 1,
  2, 4
))

vals[select]
```

Data frames have the characteristics of both lists and matrices:

When subsetting with a single index, they behave like lists and index the columns, so df[1:2] selects the first two columns.

When subsetting with two indices, they behave like matrices, so df[1:3, ] selects the first three rows (and all the columns)33.

```{r}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])

df[df$x == 2, ]

df[c(1, 3), ]

df[c("x", "z")]

df[, c("x", "z")]

# There's an important difference if you select a single 
# column: matrix subsetting simplifies by default, list 
# subsetting does not.
str(df["x"])

str(df[, "x"])
```

Subsetting a tibble with `[]` always returns a tibble
```{r}
df <- tibble::tibble(x = 1:3, y = 3:1, z = letters[1:3])

str(df["x"])

str(df[, "x"])
```

To preserve the original dimensionality, you must use `drop = FALSE`
```{r}
a <- matrix(1:4, nrow = 2)
str(a[1, ])

str(a[1, , drop = FALSE])
```

Data frames with a single column will return just the content of that column
```{r}
df <- data.frame(a = 1:2, b = 1:2)
str(df[, "a"])

str(df[, "a", drop = FALSE])
```

Factor subsetting also has a drop argument, but its meaning is rather different. It controls whether or not levels (rather than dimensions) are preserved, and it defaults to FALSE. If you find you’re using drop = TRUE a lot it’s often a sign that you should be using a character vector instead of a factor.
```{r}
z <- factor(c("a", "b"))
z[1, drop = TRUE]
```

#### 4.2 Exercises
Fix each of the following common data frame subsetting errors:
```{r}
mtcars[mtcars$cyl == 4, ]
mtcars[-c(1:4), ]
mtcars[mtcars$cyl <= 5,]
mtcars[mtcars$cyl == 4 | 6, ] # I'm not sure what the error is here?
```

Why does the following code yield five missing values? (Hint: why is it different from x[NA_real_]?)
```{r}
x <- 1:5
x[NA]
# A missing value in the index always yeilds a missing value output. So there are 5 values in the vector, all are being reported as NA
```

What does `upper.tri()` return? How does subsetting a matrix with it work? Do we need any additional subsetting rules to describe its behaviour?
```{r}
x <- outer(1:5, 1:5, FUN = "*")
x
x[upper.tri(x)]

?upper.tri # i am a bit confused about this function
```

Why does `mtcars[1:20]` return an error? How does it differ from the similar `mtcars[1:20, ]`?
```{r}
mtcars[1:20] # Not specifying rows or coluns for a 2D object
mtcars[1:20,]
```

Implement your own function that extracts the diagonal entries from a matrix (it should behave like `diag(x)` where `x` is a matrix).
```{r}
# May need further assistance with this.
```

What does `df[is.na(df)] <- 0` do? How does it work?
```{r}
df[is.na(df)] <- 0
df # it should subset all NAs in df and replace them with 0
```

### 4.3 Exercises
Brainstorm as many ways as possible to extract the third value from the `cyl` variable in the `mtcars` dataset.
```{r}
mtcars
mtcars$cyl[3]
```

Given a linear model, e.g., `mod <- lm(mpg ~ wt, data = mtcars)`, extract the residual degrees of freedom. Then extract the R squared from the model summary (`summary(mod)`)
```{r}
mod <- lm(mpg ~ wt, data = mtcars)

summary(mod)
```

#### 4.5 Exercises
How would you randomly permute the columns of a data frame? (This is an important technique in random forests.) Can you simultaneously permute the rows and columns in one step?
```{r}
df[sample(nrow(df)),]
?sample
```

How would you select a random sample of `m` rows from a data frame? What if the sample had to be contiguous (i.e., with an initial row, a final row, and every row in between)?

How could you put the columns in a data frame in alphabetical order?