---
title: "2022_10_03_chapter5"
author: "Matthew Davis"
date: "2022-10-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 5

### Quiz
What is the difference between `if` and `ifelse()`?

In the following code, what will the value of `y` be if `x` is `TRUE`? What if `x` is `FALSE`? What if `x` is `NA`?

What does `switch("x", x = , y = 2, z = 3)` return?

### 5.2.4 Exercise
1. What type of vector does each of the following calls to `ifelse()` return?
```{r}
?ifelse
class(ifelse(TRUE, 1, "no")) # Testing for a true variable creates a numeric vector
class(ifelse(FALSE, 1, "no")) # Testing for a false variable creates a character vector
class(ifelse(NA, 1, "no")) # Testing for a NA variable creates a numeric vector
```

2. Why does the following code work?
```{r}
x <- 1:10
if (length(x)) "not empty" else "empty"
# If these is any value in the x variable, the return output is "not empty"

x <- numeric()
if (length(x)) "not empty" else "empty"
# If there is no value in the x variable, then the return output is "empty"
```

### Exercise 5.3.3
1. Why does this code succeed without errors or warnings?
```{r}
x <- numeric()
out <- vector("list", length(x))
for (i in 1:length(x)) {
  out[i] <- x[i] ^ 2
}
out # It isn't because of the vector specification here. Trying out <- numeric() yielded the same result.
# Is it because the loop says to square everything in variable x, but there is nothing in variable x, so Null is still squared, which is Null
```

2. When the following code is evaluated, what can you say about the vector being iterated?
```{r}
xs <- c(1, 2, 3)
for (x in xs) {
  xs <- c(xs, x * 2)
}
xs 
```

3.What does the following code tell you about when the index is updated?
```{r}
for (i in 1:3) {
  i <- i * 2
  print(i) 
} # is the index updated at the beginning of each iteration of the loop
```

### Questions?
I am a bit confused about 5.2.1 in general when discussing the input errors.

 Can we go over Exercise 5.3.3?